---
title: "Group 8 Project"
subtitle: "PIXE analysis on Maya blue in Prehispanic and colonial mural paintings"
author: "Awet, Connor, Lachlan, Pablo,"
format: 
  html:
    toc: true
    toc-depth: 3
---

```{r}
#| include: false
library("readxl")
library(tidyverse)
library(gt)
library(GGally)
library(cowplot)
library(MVN)
library(ggcorrplot)
library(mvtnorm)
Table1_raw <- read_excel("../Data/Table1_raw.xlsx")

Table1 <- read_excel("../Data/Table1_clean.xlsx")
Table1$`17-Cl*` <- as.numeric(Table1$`17-Cl*`)
Table1$`15-P` <- as.numeric(Table1$`15-P`)
Table1$`29-Cu` <- as.numeric(Table1$`29-Cu`)
```

# Introduction

The original study by Sánchez del Río et al. (2006) investigated Maya Blue pigments in Prehispanic and colonial mural paintings from different archaeological sites. The study analysed 35 mural samples using PIXE (Particle-Induced X-ray Emission) to measure the concentrations of 17 chemical elements, ranging from sodium (Na) to lead (Pb). Principal component analysis (PCA) was then used to investigate patterns in the elemental composition of the samples. The authors reported that approximately 70% of the total variation could be explained by the first three principal components and identified structure associated with three archaeological groups: CEC, Prehispanic, and convent samples.

The aim of the present analysis is to investigate whether similar group structure can be identified in the available data and to assess the reproducibility of the main findings reported by Sánchez del Río et al. (2006). We use descriptive and graphical analyses, multivariate diagnostics, PCA, and cluster analysis to examine variation among the mural samples and evaluate whether the patterns reported in the original study can be reproduced. Particular attention is given to the PCA structure and the archaeological interpretation of the resulting groups.

# Methodology

## Data Transformation
The elemental concentration data were preprocessed before the multivariate analysis. The elements are measured on very different scales, with some reported as percentages and others as micrograms per gram (µg/g). Following Sánchez del Río et al. (2006), a base-10 logarithmic transformation was applied to reduce these large differences in scale and make the elemental concentrations easier to compare.

Some concentrations were reported in parentheses, indicating values below the limit of detection (LOD). These observations were retained because numerical estimates were still available. Values recorded as zero could not be directly log-transformed because \(\log_{10}(0)\) is undefined. Therefore, for each element, zero values were replaced by half of the smallest positive concentration observed for that element before applying the log transformation.

The resulting log-transformed concentrations were then used for the multivariate analyses, including correlation analysis, Mahalanobis-distance diagnostics, principal component analysis (PCA), and clustering.

# Results

<!-- -->

## Quantitative Descriptive Analysis

```{r}
#| include: false
### Kept this transformation in the method section.

#Non transformed table - dont use. 
# Make estimated concentrations positive
Table1 <- Table1 %>% 
  mutate(across(where(is.numeric), abs))

# Replace values recorded as zero with half of the
# smallest positive value for that element
Table1 <- Table1 %>%
  mutate(across(
    where(is.numeric),
    ~ replace(
        .x,
        is.na(.x) | .x == 0,
        min(.x[.x > 0], na.rm = TRUE) / 2
      )
  ))

#Log Table - Use as variables become comparable
Table1_log <- Table1 %>% 
  mutate(across(where(is.numeric), log10))

#Change table around to long ways for boxplots in future as current table arrangement didnt work.

#This one incase we need it for the non-transformed. 
Table1_long <- Table1 %>%
  pivot_longer(
    cols = -Elements,
    names_to = "Element_Group",     
    values_to = "Concentrations"   
  )

#Main Table we use for boxplot
Table1_long_log <- Table1_log %>%
  pivot_longer(
    cols = -Elements,
    names_to = "Element_Group",     
    values_to = "Concentrations"
  )
```

```{r}
#| echo: false
# Descriptive table of Raw data - Show why we need to transform
desc_pixe <- Table1 %>% 
  summarise(
    across(
      where(is.numeric),
      list(
        M = ~mean(.x, na.rm = TRUE),
        SD = ~sd(.x, na.rm = TRUE),
        Median = ~median(.x, na.rm = TRUE),
        Min = ~min(.x, na.rm =TRUE),
        Q1 = ~round(quantile(.x, probs = 0.25, na.rm = TRUE),2),
        IQR = ~round((quantile(.x, probs = 0.75, na.rm =TRUE) - quantile(.x, probs = 0.25, na.rm = TRUE)), 2),
        Q3 = ~round(quantile(.x, probs = 0.75, na.rm = TRUE),2),
        Max = ~max(.x, na.rm =TRUE)
      )
    )
  ) %>% 
  pivot_longer(
    everything(),
    names_to = c("Elements", "Statistics"),
    names_sep = "_",
    values_to = "Value"
  ) %>% 
  pivot_wider(
    names_from = Statistics,
    values_from = Value
  )

output_desc_pix <- desc_pixe %>% 
  gt() %>% 
  cols_label(
    Elements = "Element",
    M = md("Mean"),
    SD = md("Standard D."),
    Median = "Median",
    Min = "Minimum",
    Q1 = "1st Quartile",
    IQR = "Inter-Q Range",
    Q3 = "3rd Quartile",
    Max = "Maximum"
  ) %>% 
  fmt_number(
    columns = c(M, SD, Median, Min, Max),
    decimals = 2
  ) %>% 
  tab_header(
    title = "Descriptive Statistics of Elemental Concentrations",
    subtitle = "Non-Transformed"
  ) %>% 
  tab_options(
    table.font.size = 12
  )
output_desc_pix
```
The descriptive statistics show large differences in concentration across the elements. Calcium (Ca) occurs at relatively high concentrations in many samples, while several trace elements have much lower concentrations. Phosphorus (P) has particularly large values in some Totimehuacan samples. For example, P is 99,600 micrograms per gram in (Tot1_1) and 51,800 micrograms per gram in (Tot1_3). These values were kept because they appear in the original data, and there is not enough evidence to treat them as measurement errors. The large differences in concentration between elements also support the use of a log transformation before the multivariate analysis.
```{r}
#| echo: false
# Log Descriptives Table
desc_pixe_log <- Table1_log %>% 
  summarise(
    across(
      where(is.numeric),
      list(
        M = ~mean(.x, na.rm = TRUE),
        SD = ~sd(.x, na.rm = TRUE),
        Median = ~median(.x, na.rm = TRUE),
        Min = ~min(.x, na.rm =TRUE),
        Q1 = ~round(quantile(.x, probs = 0.25, na.rm = TRUE),2),
        IQR = ~round((quantile(.x, probs = 0.75, na.rm =TRUE) - quantile(.x, probs = 0.25, na.rm = TRUE)), 2),
        Q3 = ~round(quantile(.x, probs = 0.75, na.rm = TRUE),2),
        Max = ~max(.x, na.rm =TRUE)
      )
    )
  ) %>% 
  pivot_longer(
    everything(),
    names_to = c("Elements", "Statistics"),
    names_sep = "_",
    values_to = "Value"
  ) %>% 
  pivot_wider(
    names_from = Statistics,
    values_from = Value
  )
output_desc_pix_log <- desc_pixe_log %>% 
  gt() %>% 
  cols_label(
    Elements = "Element",
    M = md("Mean"),
    SD = md("Standard D."),
    Median = "Median",
    Min = "Minimum",
    Q1 = "1st Quartile",
    IQR = "Inter-Q Range",
    Q3 = "3rd Quartile",
    Max = "Maximum"
  ) %>% 
  fmt_number(
    columns = c(M, SD, Median, Min, Max),
    decimals = 2
  ) %>% 
  tab_header(
    title = "Descriptive Statistics of Elemental Concentrations",
    subtitle = "Log10 Transformation"
  ) %>% 
  tab_options(
    table.font.size = 12
  )
output_desc_pix_log
```
After the log transformation, the elemental concentrations are on more comparable scales, reducing the influence of the very large raw values. Differences between elements are still present, but no single variable dominates simply because of its measurement scale. This makes the transformed data more suitable for the multivariate analyses that follow.

## Graphical and Qualitative Analysis

```{r}
#| echo: false
#| message: false
#| warning: false
#Box Plot of transformed data
decp_boxplot <- ggplot(data = Table1_long_log, aes(x = Element_Group, y = Concentrations)) + 
  geom_boxplot(aes(fill = Element_Group), notch = TRUE, outlier.shape = 1.5) + 
  theme_bw() +
  theme(axis.text.x = element_text(size = 7, angle = 90, hjust = 1, vjust = 0.5), 
        legend.position = "none")
decp_boxplot
```
The log-scale boxplots show clear differences in the distributions of the elemental concentrations. The log transformation reduces the large differences in scale seen in the raw data, although some elements still show greater variability than others. Several potential outlying observations are also visible, suggesting that some mural samples have unusual elemental compositions. These observations are examined further using multivariate outlier diagnostics.

The calcium plot provides a stronger archaeological interpretation than a purely generic distribution plot. @sanchez2006pixe report that most mural samples have Ca concentrations around 30--50%, close to the 40% Ca content of calcite, supporting use of the fresco technique. The CEC samples are the conspicuous exception, with substantially lower Ca, consistent with the authors' conclusion that these murals were not prepared by the same fresco process.

## Are There Surprising Observations?

```{r}
#| echo: false
# Mahalanobis Distances 
Table1_elements_log <- Table1_log[, -1]
mu_hat <- colMeans(Table1_elements_log)
Sigma_hat <- cov(Table1_elements_log)
p <- ncol(Table1_elements_log)

dM <- mahalanobis(Table1_elements_log, center=mu_hat, cov=Sigma_hat)
upper.quantiles <- qchisq(c(.9, .95, .99), df=p)
density.at.quantiles <- dchisq(x=upper.quantiles, df=p)
cut.points <- data.frame(upper.quantiles, density.at.quantiles)
dm_plot <- ggplot(data.frame(dM), aes(x=dM)) +
  geom_histogram(aes(y=after_stat(density)), bins=nclass.FD(dM),
                 fill="white", col="black") +
  geom_rug() +
  stat_function(fun=dchisq, args = list(df=p),
                col="red", linewidth=2, alpha=.7, xlim=c(0,25)) +
  geom_segment(data=cut.points,
               aes(x=upper.quantiles, xend=upper.quantiles,
                   y=rep(0,3), yend=density.at.quantiles),
               col="blue", linewidth=2) +
  xlab("Squared Mahalanobis distances and cut points") +
  ylab("Histogram and density")

Table1_elements_log$dM <- dM
Table1_elements_log$suprise <- cut(Table1_elements_log$dM,
                                   breaks = c(0, upper.quantiles, Inf),
                                   labels=c("typical", "somewhat", "Suprising", "very"))

dm_results <- data.frame(
  Sample = Table1_log[[1]],
  dM = dM,
  surprise = Table1_elements_log$suprise
)

dm_results <- dm_results %>% 
  dplyr::arrange(desc(dM)) %>% 
  head(5)
dm_plot
table(Table1_elements_log$suprise)
dm_results
```
The squared Mahalanobis distances were used to identify samples with unusual combinations of elemental concentrations. Most observations have relatively smaller Mahalanobis distance, while a smaller number have larger distance and may be considered potentially unusual. The 90%, 95%, and 99% chi-square cut-offs provide increasing levels of evidence for identifying unusual multivariate observations. These samples are retained in the analysis because an unusual observation is not necessarily an error; it may represent genuine differences in mural composition.

```{r}
#| echo: false
#| message: false
#| warning: false
#Had problems with the pairs plot from lectures mainly with the contours and scatter-points
#Two fix I built the two types of plots seperately so I can change the sizes and bins
#Then added them in seperately in the ggpairs command like done in the lectures
#The main problem was that they become squished when using the inbuilt command due to
#their being 17 variables. So doing them seperately helped fix that

surprise_points <- function(data, mapping, ...){
  ggplot(data = data, mapping = mapping) +
    geom_point(
      aes(colour = suprise),
      alpha = 0.5)}
upper_contour <- function(data, mapping, ...) {
  ggplot(data = data, mapping = mapping) +
    stat_density_2d(
      colour = "blue",
      linewidth = 0.3,
      bins = 3)}

pairs <- ggpairs(data = Table1_elements_log,
  columns = 1:17,
  lower = list(continuous = surprise_points),
  upper = list(continuous = upper_contour),
  diag = list(continuous = wrap(
      "densityDiag",
      colour = "black",
      fill = "grey"))) +
  scale_color_manual(
    values = c(
      "lightgray",
      "green",
      "blue",
      "red")) +
  theme(axis.text = element_blank(),
    axis.ticks = element_blank(),
    strip.text = element_text(size = 7))

# Legend for the pairs plot 
#Tried adding the main part to the pairs plot code itself but wouldnt show so 
#doing it seperately using cowplot. 
legend_plot <- ggplot(
  Table1_elements_log,
  aes(x = suprise, y = 1, colour = suprise)
) +
  geom_point() +
  scale_color_manual(
    values = c(
      "typical" = "lightgray",
      "somewhat" = "green",
      "Suprising" = "blue",
      "very" = "red"
    ),
    labels = c(
      "typical" = "Typical (<90%)",
      "somewhat" = "Somewhat surprising (90–95%)",
      "Suprising" = "Surprising (95–99%)",
      "very" = "Very surprising (>99%)"
    ),
    name = "Mahalanobis classification",
    drop = FALSE
  ) +
  guides(
    colour = guide_legend(
      nrow = 1,
      override.aes = list(alpha = 1, size = 3)
    )
  ) +
  theme_void() +
  theme(
    legend.position = "right",
    legend.box.margin = margin(5,10,5,10)
  )
#cowplot adds the legend that we made separately to the pairs plot
#Very annoying work but none the less like the result
legend <- cowplot::get_legend(legend_plot)
#Something that cowplot can use as ggpairs isnt a ggplot2 ggplot
pairs_grob <- GGally::ggmatrix_gtable(pairs)

final_pairs <- cowplot::plot_grid(
  pairs_grob,
  legend,
  ncol = 1,
  rel_heights = c(1, 0.14)
)
#Takes about 15 seconds to load.
final_pairs
```
The pairs plot shows the relationships among the 17 log-transformed elements. Most samples are classified as typical or somewhat surprising based on their Mahalanobis distances, while only a few are classified as more unusual. The direction and strength of the relationships vary between pairs of elements. These patterns help identify potentially unusual samples and highlight element pairs that are useful for further investigation.

## Can you assume that the data follows a multivariate normal distribution?

```{r}
#| echo: false
#We use the DM^2 values and compare them to the chi-square quantiles.
qq_elements <- Table1_elements_log[, -c(18, 19)]
n <- nrow(qq_elements)
p <- ncol(qq_elements)

chi_quantiles <- qchisq(
  ppoints(n),
  df = p
)

qq_data <- data.frame(
  Theoretical = qchisq(ppoints(n), df = p),
  Observed = sort(dM)
)

#Chi-square QQ-plot
normality_plot <- ggplot(qq_data,aes(x = Theoretical, y = Observed)) +
  geom_point(size = 2) +
  geom_abline(intercept = 0,
    slope = 1,
    linetype = "dashed",
    color = "red") +
  labs(x = "Theoretical chi-square(17) quantiles",
    y = "Ordered dm^2",
    title = "Multivariate Q-Q Plot"
  ) +
  theme_minimal()
normality_plot

#Mardia Multivariate Normal Test and individual anderson-darling tests of normality
mvn_result <- mvn(
  data = qq_elements,
  mvn_test = "mardia"
)
mvn_result$multivariate_normality
mvn_result$univariate_normality
```
The multivariate normality assumption is not fully satisfied. Mardia's skewness test is significant (p = 0.003), providing evidence against multivariate normality, although Mardia's kurtosis test is not significant (p = 0.664). The multivariate Q-Q plot is approximately linear through much of the central region but shows some departures from the reference line. Therefore, results that rely on multivariate normality, particularly the chi-square interpretation of Mahalanobis distances, should be interpreted with caution.
## Can we find groups in this data

```{r}
#| echo: false
#Correlation Heatmap: Pablo
#Fixed Issue with plot, changed to [,-1] as [,-(1:2)] includes columns 1 and 2, which removes 11-Na* from the plot
#[,-1] removes only the elements column as intended
ggcorrplot(cor(Table1_log[,-1]),
           method = "circle",
           hc.order = TRUE,
           type = "lower")
```

The correlogram shows the direction and strength of the relationships among the log-transformed elemental concentrations. Red circles represent positive correlations and blue circles represent negative correlations, while larger and darker circles indicate stronger relationships. Several element pairs show clear associations, suggesting that some elemental concentrations tend to vary together across the mural samples.

To examine these patterns more closely, two contrasting relationships from the correlogram are considered below.

### Relationship between 12-Mg\* and 14-Si\*

```{r}
#| include: false
#Look at two specific relationships too show off and talk about. 
mu <- colMeans(Table1_log[,-1])
Sigma <- var(Table1_log[,-1])
rho <- cor(Table1_log[,-1])

#Means
(mu_mg_si <- mu[c("12-Mg*", "14-Si*")])
#Covariance Matrix
(Sigma_mg_si <- Sigma[c("12-Mg*", "14-Si*"), c("12-Mg*", "14-Si*")])
#Correlation Matrix
(rho_mg_si <- rho[c("12-Mg*", "14-Si*"), c("12-Mg*", "14-Si*")])
```

```{r}
#| echo: false
#| warning: false
#| fig-width: 6
#| fig-height: 6
#| fig-align: center
#Plot of 
n <- 100
x <- seq(mu_mg_si[1]-4*sqrt(Sigma_mg_si[1,1]),
         mu_mg_si[1]+4*sqrt(Sigma_mg_si[1,1]), len=n)
y <- seq(mu_mg_si[2]-4*sqrt(Sigma_mg_si[2,2]), 
         mu_mg_si[2]+4*sqrt(Sigma_mg_si[2,2]), len=n)
xy <- expand.grid(x=x, y=y)

q.samp <- cbind(xy, prob=dmvnorm(xy, mean = mu_mg_si, sigma = Sigma_mg_si))

ggplot(q.samp, aes(x=y, y=x, z=prob)) + 
  geom_point(data=Table1_log, aes(y=`12-Mg*`, x=`14-Si*`, z=0)) +
  geom_contour(bins=5, col="red") +
  ylab("12-Mg*") + xlab("14-Si*") + ggtitle("Bivariate Scatterplot of 12-Mg* and 14-Si*", "With Two-Dimensional Density Contours") +
  coord_fixed(
    ylim = c(mu_mg_si[1]-4*sqrt(Sigma_mg_si[1,1]), mu_mg_si[1]+4*sqrt(Sigma_mg_si[1,1])),
    xlim = c(mu_mg_si[2]-4*sqrt(Sigma_mg_si[2,2]), mu_mg_si[2]+4*sqrt(Sigma_mg_si[2,2]))
    ) +
  theme(axis.text.x = element_text(angle = 45))
```

The bivariate scatterplot shows a clear positive association between the log-transformed concentrations of Mg and Si. The upward-sloping density contours support this pattern: samples with higher Si concentrations generally also have higher Mg concentrations. The relatively concentrated contours suggest that this relationship is fairly strong compared with some of the other element pairs.

### Relationship between 22-Ti and 17-CI\*

```{r}
#| include: false
mu <- colMeans(Table1_log[,-1])
Sigma <- var(Table1_log[,-1])
rho <- cor(Table1_log[,-1])

#Means
(mu_mg_si <- mu[c("22-Ti", "17-Cl*")])
#Covariance Matrix
(Sigma_mg_si <- Sigma[c("22-Ti", "17-Cl*"), c("22-Ti", "17-Cl*")])
#Correlation Matrix
(rho_mg_si <- rho[c("22-Ti", "17-Cl*"), c("22-Ti", "17-Cl*")])
```

```{r}
#| echo: false
#| warning: false
#| fig-width: 6
#| fig-height: 6
#| fig-align: center
#Plot of 
n <- 100
x <- seq(mu_mg_si[1]-4*sqrt(Sigma_mg_si[1,1]),
         mu_mg_si[1]+4*sqrt(Sigma_mg_si[1,1]), len=n)
y <- seq(mu_mg_si[2]-4*sqrt(Sigma_mg_si[2,2]), 
         mu_mg_si[2]+4*sqrt(Sigma_mg_si[2,2]), len=n)
xy <- expand.grid(x=x, y=y)

q.samp <- cbind(xy, prob=dmvnorm(xy, mean = mu_mg_si, sigma = Sigma_mg_si))

ggplot(q.samp, aes(x=y, y=x, z=prob)) + 
  geom_point(data=Table1_log, aes(y=`22-Ti`, x=`17-Cl*`, z=0)) +
  geom_contour(bins=5, col="red") +
  ylab("22-Ti") + xlab("17-Cl*") + ggtitle("Bivariate Scatterplot of 22-Ti and 17-Cl*", "With Two-Dimensional Density Contours") +
  coord_fixed(
    ylim = c(mu_mg_si[1]-4*sqrt(Sigma_mg_si[1,1]), mu_mg_si[1]+4*sqrt(Sigma_mg_si[1,1])),
    xlim = c(mu_mg_si[2]-4*sqrt(Sigma_mg_si[2,2]), mu_mg_si[2]+4*sqrt(Sigma_mg_si[2,2]))
    ) +
  theme(axis.text.x = element_text(angle = 45))
```

As we can see from the Bivariate Scatter plot of 22-Ti and 17-Cl, there is a negative association between the log-transformed concentrations of the two elements. the downward oreientation of the density contours suggest that samples with higher Cl concentrations generally have lower Ti concentrations. However, this relationship is relatively weaker then the relationship shown between 12-Mg and 14-Si as indicated by the dispersion of points and broader contour lines.

## PCA

We can now conduct a Principle component analysis of the data to find how many groups are in the data. The Covariance matrix was used instead of the correlation matrix to keep inline with the original studies PCA.

```{r}
#| echo: false
#PCA
PCA_log <- prcomp(Table1_log[,-1], 
                  center=TRUE, 
                  scale=FALSE)
plot(PCA_log, type="l")

PCA_variance <- data.frame(
  PC = paste0("PC", 1:length(PCA_log$sdev)),
  Eigenvalue = round(PCA_log$sdev^2, 3),
  `Proportion of Variance (%)` =
    round(100 * PCA_log$sdev^2 / sum(PCA_log$sdev^2), 2),
  `Cumulative Variance (%)` =
    round(100 * cumsum(PCA_log$sdev^2 / sum(PCA_log$sdev^2)), 2)
)

knitr::kable(
  PCA_variance,
  caption = "Variance explained by each principal component"
)
```

The first three principal components explained 77.08% of the total variance, with PC1, PC2, and PC3 accounting for 34.02%, 25.00%, and 18.05%, respectively. These results are broadly consistent with the original study, which reported 34.88%, 24.95%, and 11.32% for the first three components (Sanchez del Rio et al, 2006). The proportions of theb  variance by the first components are particularly similar to those reported in the in the orginal study. However the third principle component in our analysis is noticeably larger then the percentage of variance in the original study.

## Can we reproduce Fig.2

```{r}
#| echo: false
#Cluster Analysis
```

# Discussion

Answering the questions from the document and any other conclusions we have from the analysis

1.  Can you find groups in these data?
2.  Can you reproduce Fig. 2? Do you agree with the conclusions the authors draw about the “zones”?

# References {style="text-align: center;"}

> Andonie, R., and I. Dzitac. 2010. “How to Write a Good Paper in Computer Science and How Will It Be Measured by ISI Web of Knowledge.” International Journal of Computers, Communications & Control V (4): 432–44. https://doi.org/10.15837/ijccc.2010.4.2493.
>
> Frery, A. C., L. Gomez, and A. C. Medeiros. 2020. “A Badging System for Reproducibility and Replicability in Remote Sensing Research.” IEEE Journal of Selected Topics on Applied Earth Observations and Remote Sensing 13: 4988–95. https://doi.org/10.1109/JSTARS.2020.3019 418.
>
> Sánchez del Río, M., Martinetto, P., Solís, C. & Reyes-Valerio, C. (2006), ‘PIXE analysis on Maya blue in Prehispanic and colonial mural paintings’, Nuclear Instruments and Methods in Physics Research Section B: Beam Interactions with Materials and Atoms 249(1–2), 628–632.
