---
title: "Group 8 Project"
subtitle: "PIXE analysis on Maya blue in Prehispanic and colonial mural paintings"
author: "Awet, Connor, Lachlan, Pablo,"
format: 
  html:
    toc: true
    toc-depth: 3
---

```{r}
#| include: false
library("readxl")
library(tidyverse)
library(gt)
library(GGally)
library(cowplot)
library(MVN)
Table1_raw <- read_excel("../Data/Table1_raw.xlsx")

Table1 <- read_excel("../Data/Table1_clean.xlsx")
Table1$`17-Cl*` <- as.numeric(Table1$`17-Cl*`)
Table1$`15-P` <- as.numeric(Table1$`15-P`)
Table1$`29-Cu` <- as.numeric(Table1$`29-Cu`)
```

# Introduction

The original article, PIXE analysis on Maya blue in Prehispanic and colonial mural paintings, by Sanchez del Rio et al (2006), sampled, analysed and grouped using PCA on murals containing Maya Blue across different archaeological sites. Across the sites they collected 35 samples from murals and either the elements concentration percent or the ratio of micro gram to gram ratio of that element in the mural. From these murals they sampled 17 different chemical elements from sodium to lead. They found with their PCA that 70% of the total variance in the sample population can be described using only the first three principle components and was similar to what was observed in other art forms. The three zones found were CEC, Prehispanic, and convents. with the first clustering in a small group separate from the rest of the samples.

The aim of this analysis is too find groups within the data and reproduce what was found in the original paper. The reason for this is to support the conclusions made by Sanchez del rio et al (2006) about the 3 zones found in the data during their analysis. By doing this we can reassert the claims made about the differences between the different types of murals and their distinct zones of style. This report will contain both a quantitative and qualitative analysis of the data, as well as PCA and cluster analysis to explore the possible groups in the data.

# Methodology

## Data Transformation 

There are a few issues with the original data set in Table 1 that need to be addressed before performing the analysis.

1. Different concentration ranges: The original data contain elemental concentrations reported either as percentages or as µg/g (micrograms per gram). Because these measurements have substantially different concentration ranges, a \(\log_{10}\) transformation is applied to the concentrations. This follows the procedure used in the original paper, where the authors state that the logarithm of the concentrations was used to avoid large discrepancies in concentration range.
2. Values below the detection limit: Some concentrations in Table 1 are reported in parentheses ( ). According to the original paper, these values are below the limit of detection (LOD). Since numerical estimates are still provided, these estimated concentrations are retained for the analysis and the parentheses are removed.
3. ### Replace with Awets transformation of NA's and 0's 


# Results
<!-- -->

## Quantitative Descriptive Analysis

```{r}
#| include: false
### Kept this transformation in the method section.

#Non transformed table - dont use. 
# Make estimated concentrations positive
Table1 <- Table1 %>% 
  mutate(across(where(is.numeric), abs))

# Replace values recorded as zero with half of the
# smallest positive value for that element
Table1 <- Table1 %>%
  mutate(across(where(is.numeric),
      ~ {replacement <- min(.x[.x > 0], na.rm = TRUE) / 2
        ifelse(is.na(.x) | .x == 0,replacement, .x)}))

#Log Table - Use as variables become comparable
Table1_log <- Table1 %>% 
  mutate(across(where(is.numeric), log10))

#Change table around to long ways for boxplots in future as current table arrangement didnt work.

#This one incase we need it for the non-transformed. 
Table1_long <- Table1 %>%
  pivot_longer(
    cols = -Elements,
    names_to = "Element_Group",     
    values_to = "Concentrations"   
  )

#Main Table we use for boxplot
Table1_long_log <- Table1_log %>%
  pivot_longer(
    cols = -Elements,
    names_to = "Element_Group",     
    values_to = "Concentrations"
  )
```

```{r}
#| echo: false
# Descriptive table of Raw data - Show why we need to transform
desc_pixe <- Table1 %>% 
  summarise(
    across(
      where(is.numeric),
      list(
        M = ~mean(.x, na.rm = TRUE),
        SD = ~sd(.x, na.rm = TRUE),
        Median = ~median(.x, na.rm = TRUE),
        Min = ~min(.x, na.rm =TRUE),
        Max = ~max(.x, na.rm =TRUE)
      )
    )
  ) %>% 
  pivot_longer(
    everything(),
    names_to = c("Elements", "Statistics"),
    names_sep = "_",
    values_to = "Value"
  ) %>% 
  pivot_wider(
    names_from = Statistics,
    values_from = Value
  )

output_desc_pix <- desc_pixe %>% 
  gt() %>% 
  cols_label(
    Elements = "Element",
    M = md("Mean"),
    SD = md("*SD*"),
    Median = "Median",
    Min = "Minimum",
    Max = "Maximum"
  ) %>% 
  fmt_number(
    columns = c(M, SD, Median, Min, Max),
    decimals = 2
  ) %>% 
  tab_header(
    title = "Descriptive Statistics of Elemental Concentrations",
    subtitle = "Non-Transformed"
  ) %>% 
  tab_options(
    table.font.size = 12
  )
output_desc_pix
```

```{r}
#| echo: false
# Log Descriptives Table
desc_pixe_log <- Table1_log %>% 
  summarise(
    across(
      where(is.numeric),
      list(
        M = ~mean(.x, na.rm = TRUE),
        SD = ~sd(.x, na.rm = TRUE),
        Median = ~median(.x, na.rm = TRUE),
        Min = ~min(.x, na.rm =TRUE),
        Max = ~max(.x, na.rm =TRUE)
      )
    )
  ) %>% 
  pivot_longer(
    everything(),
    names_to = c("Elements", "Statistics"),
    names_sep = "_",
    values_to = "Value"
  ) %>% 
  pivot_wider(
    names_from = Statistics,
    values_from = Value
  )
output_desc_pix_log <- desc_pixe_log %>% 
  gt() %>% 
  cols_label(
    Elements = "Element",
    M = md("Mean"),
    SD = md("*SD*"),
    Median = "Median",
    Min = "Minimum",
    Max = "Maximum"
  ) %>% 
  fmt_number(
    columns = c(M, SD, Median, Min, Max),
    decimals = 2
  ) %>% 
  tab_header(
    title = "Descriptive Statistics of Elemental Concentrations",
    subtitle = "Log10 Transformation"
  ) %>% 
  tab_options(
    table.font.size = 12
  )
output_desc_pix_log
```

## Graphical and Qualitative Analysis

```{r}
#| echo: false
#| message: false
#| warning: false
#Box Plot of transformed data
decp_boxplot <- ggplot(data = Table1_long_log, aes(x = Element_Group, y = Concentrations)) + 
  geom_boxplot(aes(fill = Element_Group), notch = TRUE, outlier.shape = 1.5) + 
  theme_bw() +
  theme(axis.text.x = element_text(size = 7, angle = 90, hjust = 1, vjust = 0.5), 
        legend.position = "none")
decp_boxplot
```

## Are There Surprising Observations?

```{r}
#| echo: false
# Mahalanobis Distances 
Table1_elements_log <- Table1_log[, -1]
mu_hat <- colMeans(Table1_elements_log)
Sigma_hat <- cov(Table1_elements_log)
p <- ncol(Table1_elements_log)

dM <- mahalanobis(Table1_elements_log, center=mu_hat, cov=Sigma_hat)
upper.quantiles <- qchisq(c(.9, .95, .99), df=p)
density.at.quantiles <- dchisq(x=upper.quantiles, df=p)
cut.points <- data.frame(upper.quantiles, density.at.quantiles)
dm_plot <- ggplot(data.frame(dM), aes(x=dM)) +
  geom_histogram(aes(y=after_stat(density)), bins=nclass.FD(dM),
                 fill="white", col="black") +
  geom_rug() +
  stat_function(fun=dchisq, args = list(df=p),
                col="red", linewidth=2, alpha=.7, xlim=c(0,25)) +
  geom_segment(data=cut.points,
               aes(x=upper.quantiles, xend=upper.quantiles,
                   y=rep(0,3), yend=density.at.quantiles),
               col="blue", linewidth=2) +
  xlab("Squared Mahalanobis distances and cut points") +
  ylab("Histogram and density")

Table1_elements_log$dM <- dM
Table1_elements_log$suprise <- cut(Table1_elements_log$dM,
                                   breaks = c(0, upper.quantiles, Inf),
                                   labels=c("typical", "somewhat", "Suprising", "very"))

dm_results <- data.frame(
  Sample = Table1_log[[1]],
  dM = dM,
  surprise = Table1_elements_log$suprise
)

dm_results <- dm_results %>% 
  dplyr::arrange(desc(dM)) %>% 
  head(5)
dm_plot
table(Table1_elements_log$suprise)
dm_results
```


```{r}
#| echo: false
#| message: false
#| warning: false
#Had problems with the pairs plot from lectures mainly with the contours and scatter-points
#Two fix I built the two types of plots seperately so I can change the sizes and bins
#Then added them in seperately in the ggpairs command like done in the lectures
#The main problem was that they become squished when using the inbuilt command due to
#their being 17 variables. So doing them seperately helped fix that

surprise_points <- function(data, mapping, ...){
  ggplot(data = data, mapping = mapping) +
    geom_point(
      aes(colour = suprise),
      alpha = 0.5)}
upper_contour <- function(data, mapping, ...) {
  ggplot(data = data, mapping = mapping) +
    stat_density_2d(
      colour = "blue",
      linewidth = 0.3,
      bins = 3)}

pairs <- ggpairs(data = Table1_elements_log,
  columns = 1:17,
  lower = list(continuous = surprise_points),
  upper = list(continuous = upper_contour),
  diag = list(continuous = wrap(
      "densityDiag",
      colour = "black",
      fill = "grey"))) +
  scale_color_manual(
    values = c(
      "lightgray",
      "green",
      "blue",
      "red")) +
  theme(axis.text = element_blank(),
    axis.ticks = element_blank(),
    strip.text = element_text(size = 7))

# Legend for the pairs plot 
#Tried adding the main part to the pairs plot code itself but wouldnt show so 
#doing it seperately using cowplot. 
legend_plot <- ggplot(
  Table1_elements_log,
  aes(x = suprise, y = 1, colour = suprise)
) +
  geom_point() +
  scale_color_manual(
    values = c(
      "typical" = "lightgray",
      "somewhat" = "green",
      "Suprising" = "blue",
      "very" = "red"
    ),
    labels = c(
      "typical" = "Typical (<90%)",
      "somewhat" = "Somewhat surprising (90–95%)",
      "Suprising" = "Surprising (95–99%)",
      "very" = "Very surprising (>99%)"
    ),
    name = "Mahalanobis classification",
    drop = FALSE
  ) +
  guides(
    colour = guide_legend(
      nrow = 1,
      override.aes = list(alpha = 1, size = 3)
    )
  ) +
  theme_void() +
  theme(
    legend.position = "right",
    legend.box.margin = margin(5,10,5,10)
  )
#cowplot adds the legend that we made separately to the pairs plot
#Very annoying work but none the less like the result
legend <- cowplot::get_legend(legend_plot)
#Something that cowplot can use as ggpairs isnt a ggplot2 ggplot
pairs_grob <- GGally::ggmatrix_gtable(pairs)

final_pairs <- cowplot::plot_grid(
  pairs_grob,
  legend,
  ncol = 1,
  rel_heights = c(1, 0.14)
)
#Takes about 15 seconds to load.
final_pairs
```

## Can you assume that the data follows a multivariate normal distribution?

```{r}
#| echo: false
#We use the DM^2 values and compare them to the chi-square quantiles.
qq_elements <- Table1_elements_log[, -c(18, 19)]
n <- nrow(qq_elements)
p <- ncol(qq_elements)

chi_quantiles <- qchisq(
  ppoints(n),
  df = p
)

qq_data <- data.frame(
  Theoretical = qchisq(ppoints(n), df = p),
  Observed = sort(dM)
)

#Plot
normality_plot <- ggplot(qq_data,aes(x = Theoretical, y = Observed)) +
  geom_point(size = 2) +
  geom_abline(intercept = 0,
    slope = 1,
    linetype = "dashed",
    color = "red") +
  labs(x = "Theoretical chi-square(17) quantiles",
    y = "Ordered dm^2",
    title = "Multivariate Q-Q Plot"
  ) +
  theme_minimal()
normality_plot

mvn_result <- mvn(
  data = qq_elements,
  mvn_test = "mardia"
)
mvn_result$multivariate_normality
mvn_result$univariate_normality
```


## Can we find groups in this data

## Can we reproduce Fig.2





# Discussion

Answering the questions from the document and any other conclusions we have from the analysis

1.  Can you find groups in these data?
2.  Can you reproduce Fig. 2? Do you agree with the conclusions the authors draw about the “zones”?

# References {style="text-align: center;"}

> Andonie, R., and I. Dzitac. 2010. “How to Write a Good Paper in Computer Science and How Will It Be Measured by ISI Web of Knowledge.” International Journal of Computers, Communications & Control V (4): 432–44. https://doi.org/10.15837/ijccc.2010.4.2493.
>
> Frery, A. C., L. Gomez, and A. C. Medeiros. 2020. “A Badging System for Reproducibility and Replicability in Remote Sensing Research.” IEEE Journal of Selected Topics on Applied Earth Observations and Remote Sensing 13: 4988–95. https://doi.org/10.1109/JSTARS.2020.3019 418.
>
> Sánchez del Río, M., Martinetto, P., Solís, C. & Reyes-Valerio, C. (2006), ‘PIXE analysis on Maya blue in Prehispanic and colonial mural paintings’, Nuclear Instruments and Methods in Physics Research Section B: Beam Interactions with Materials and Atoms 249(1–2), 628–632.
