---
title: "Group 8 Project"
subtitle: "PIXE analysis on Maya blue in Prehispanic and colonial mural paintings"
author: "Awet, Connor, Lachlan, Pablo,"
format: 
  html:
    toc: true
    toc-depth: 3
---

```{r}
#| include: false
library("readxl")
library(tidyverse)
library(gt)
library(GGally)
library(cowplot)
library(MVN)
library(ggcorrplot)
library(mvtnorm)
library(knitr)
Table1_raw <- read_excel("../Data/Table1_raw.xlsx")

Table1 <- read_excel("../Data/Table1_clean.xlsx")
Table1$`17-Cl*` <- as.numeric(Table1$`17-Cl*`)
Table1$`15-P` <- as.numeric(Table1$`15-P`)
Table1$`29-Cu` <- as.numeric(Table1$`29-Cu`)
```

# Introduction

The original study by Sánchez del Río et al. (2006) investigated Maya Blue pigments in Prehispanic and colonial mural paintings from different archaeological sites. The study analysed 35 mural samples using PIXE (Particle-Induced X-ray Emission) to measure the concentrations of 17 chemical elements, ranging from sodium (Na) to lead (Pb). Principal component analysis (PCA) was then used to investigate patterns in the elemental composition of the samples. The authors reported that approximately 70% of the total variation could be explained by the first three principal components and identified structure associated with three archaeological groups: CEC, Prehispanic, and convent samples.

The aim of the present analysis is to investigate whether similar group structure can be identified in the available data and to assess the reproducibility of the main findings reported by Sánchez del Río et al. (2006). We use descriptive and graphical analyses, multivariate diagnostics, PCA, and cluster analysis to examine variation among the mural samples and evaluate whether the patterns reported in the original study can be reproduced. Particular attention is given to the PCA structure and the archaeological interpretation of the resulting groups.

# Methodology

## Data
```{r}
#| echo: false
kable(Table1)
```

Data for this analysis contains 35 samples of different types of murals from across regions of southern mexico (Sanchez del Rio et al, 2006). This data can be found in Table 1. of Sanchez del Rio et al's paper from 2006. 

## Data Transformation
The elemental concentration data were preprocessed before the multivariate analysis. The elements are measured on very different scales, with some reported as percentages and others as micrograms per gram (µg/g). Following Sánchez del Río et al. (2006), a base-10 logarithmic transformation was applied to reduce these large differences in scale and make the elemental concentrations easier to compare.Values below the detection limit: Some concentrations in Table 1 are reported in parentheses ( ). According to the original paper, these values are below the limit of detection (LOD). Since numerical estimates are still provided, these estimated concentrations are retained for the analysis and the parentheses are removed. Lastly we conducted a transformation of both NA and 0 values found in the data. Sanchez del Rio et al (2006) described that values that were given as 0's arent treated as 0's, and were instead concentrations that were below the detection limit, and as such were instead very small concentrations of elements. Moreover we cannot do a (\log\_{10}) transformation of the data if observations contain 0's. This leaves us to therefore transform these 0's into estimates to both do this transformation, and not remove concentrations that are there but very small. Therefore we run a function that runs through each column that is numeric, and then filters out any values that are less than or equal to zero, ignores NAs (na.rm = TRUE), finds the smallest remaining positive number, and divides it by two. Finally it checks every cell in that column. If a cell is either NA or exactly 0, it substitutes it with the replacement value.

The resulting log-transformed concentrations were then used for the multivariate analyses, including correlation analysis, Mahalanobis-distance diagnostics, principal component analysis (PCA), and clustering.


# Results

<!-- -->

## Quantitative Descriptive Analysis

```{r}
#| include: false
### Kept this transformation in the method section.

#Non transformed table - dont use. 
# Make estimated concentrations positive
Table1 <- Table1 %>% 
  mutate(across(where(is.numeric), abs))

# Replace values recorded as zero with half of the
# smallest positive value for that element
Table1 <- Table1 %>%
  mutate(across(
    where(is.numeric),
    ~ replace(
        .x,
        is.na(.x) | .x == 0,
        min(.x[.x > 0], na.rm = TRUE) / 2
      )
  ))

#Log Table - Use as variables become comparable
Table1_log <- Table1 %>% 
  mutate(across(where(is.numeric), log10))

#Change table around to long ways for boxplots in future as current table arrangement didnt work.

#This one incase we need it for the non-transformed. 
Table1_long <- Table1 %>%
  pivot_longer(
    cols = -Elements,
    names_to = "Element_Group",     
    values_to = "Concentrations"   
  )

#Main Table we use for boxplot
Table1_long_log <- Table1_log %>%
  pivot_longer(
    cols = -Elements,
    names_to = "Element_Group",     
    values_to = "Concentrations"
  )
```

```{r}
#| echo: false
# Descriptive table of Raw data - Show why we need to transform
desc_pixe <- Table1 %>% 
  summarise(
    across(
      where(is.numeric),
      list(
        M = ~mean(.x, na.rm = TRUE),
        SD = ~sd(.x, na.rm = TRUE),
        Median = ~median(.x, na.rm = TRUE),
        Min = ~min(.x, na.rm =TRUE),
        Q1 = ~round(quantile(.x, probs = 0.25, na.rm = TRUE),2),
        IQR = ~round((quantile(.x, probs = 0.75, na.rm =TRUE) - quantile(.x, probs = 0.25, na.rm = TRUE)), 2),
        Q3 = ~round(quantile(.x, probs = 0.75, na.rm = TRUE),2),
        Max = ~max(.x, na.rm =TRUE)
      )
    )
  ) %>% 
  pivot_longer(
    everything(),
    names_to = c("Elements", "Statistics"),
    names_sep = "_",
    values_to = "Value"
  ) %>% 
  pivot_wider(
    names_from = Statistics,
    values_from = Value
  )

output_desc_pix <- desc_pixe %>% 
  gt() %>% 
  cols_label(
    Elements = "Element",
    M = md("Mean"),
    SD = md("Standard D."),
    Median = "Median",
    Min = "Minimum",
    Q1 = "1st Quartile",
    IQR = "Inter-Q Range",
    Q3 = "3rd Quartile",
    Max = "Maximum"
  ) %>% 
  fmt_number(
    columns = c(M, SD, Median, Min, Max),
    decimals = 2
  ) %>% 
  tab_header(
    title = "Descriptive Statistics of Elemental Concentrations",
    subtitle = "Non-Transformed"
  ) %>% 
  tab_options(
    table.font.size = 12
  )
output_desc_pix
```

The descriptive statistics show large differences in concentration across the elements. Calcium (Ca) occurs at relatively high concentrations in many samples, while several trace elements have much lower concentrations. Phosphorus (P) has particularly large values in some Totimehuacan samples. For example, P is 99,600 micrograms per gram in (Tot1_1) and 51,800 micrograms per gram in (Tot1_3). These values were kept because they appear in the original data, and there is not enough evidence to treat them as measurement errors. The large differences in concentration between elements also support the use of a log transformation before the multivariate analysis.

```{r}
#| echo: false
# Log Descriptives Table
desc_pixe_log <- Table1_log %>% 
  summarise(
    across(
      where(is.numeric),
      list(
        M = ~mean(.x, na.rm = TRUE),
        SD = ~sd(.x, na.rm = TRUE),
        Median = ~median(.x, na.rm = TRUE),
        Min = ~min(.x, na.rm =TRUE),
        Q1 = ~round(quantile(.x, probs = 0.25, na.rm = TRUE),2),
        IQR = ~round((quantile(.x, probs = 0.75, na.rm =TRUE) - quantile(.x, probs = 0.25, na.rm = TRUE)), 2),
        Q3 = ~round(quantile(.x, probs = 0.75, na.rm = TRUE),2),
        Max = ~max(.x, na.rm =TRUE)
      )
    )
  ) %>% 
  pivot_longer(
    everything(),
    names_to = c("Elements", "Statistics"),
    names_sep = "_",
    values_to = "Value"
  ) %>% 
  pivot_wider(
    names_from = Statistics,
    values_from = Value
  )
output_desc_pix_log <- desc_pixe_log %>% 
  gt() %>% 
  cols_label(
    Elements = "Element",
    M = md("Mean"),
    SD = md("Standard D."),
    Median = "Median",
    Min = "Minimum",
    Q1 = "1st Quartile",
    IQR = "Inter-Q Range",
    Q3 = "3rd Quartile",
    Max = "Maximum"
  ) %>% 
  fmt_number(
    columns = c(M, SD, Median, Min, Max),
    decimals = 2
  ) %>% 
  tab_header(
    title = "Descriptive Statistics of Elemental Concentrations",
    subtitle = "Log10 Transformation"
  ) %>% 
  tab_options(
    table.font.size = 12
  )
output_desc_pix_log
```
After the log transformation, the elemental concentrations are on more comparable scales, reducing the influence of the very large raw values. Differences between elements are still present, but no single variable dominates simply because of its measurement scale. This makes the transformed data more suitable for the multivariate analyses that follow.

## Graphical and Qualitative Analysis

```{r}
#| echo: false
#| message: false
#| warning: false
#Box Plot of transformed data
decp_boxplot <- ggplot(data = Table1_long_log, aes(x = Element_Group, y = Concentrations)) + 
  geom_boxplot(aes(fill = Element_Group), notch = TRUE, outlier.shape = 1.5) + 
  theme_bw() +
  theme(axis.text.x = element_text(size = 7, angle = 90, hjust = 1, vjust = 0.5), 
        legend.position = "none")
decp_boxplot
```
The log-scale box plots show that transformation reduces, but does not remove, the pronounced differences in spread among elements. The trace variables remain particularly heterogeneous, while several major matrix components are more concentrated. We can see that Chlorine has the lowest mean of the 17 elements Titanium has the largest median. A number of the elements have outliers including Manganese with 4, Chlorine with 3, Phosphorous, Calcium and Nickel with 2, and finally Sodium and Aluminium having 1. This can be expected as certain types of murals Notable due to the transformation of the data, and the large amount of NA's found in the samples for concentrations of Phosphorous, it contains a very large inter quartile range with, as well as having its median, 1st quatile, a minimum value all being 1.93. 
## Are There Surprising Observations?

```{r}
#| echo: false
# Mahalanobis Distances 
Table1_elements_log <- Table1_log[, -1]
mu_hat <- colMeans(Table1_elements_log)
Sigma_hat <- cov(Table1_elements_log)
p <- ncol(Table1_elements_log)

dM <- mahalanobis(Table1_elements_log, center=mu_hat, cov=Sigma_hat)
upper.quantiles <- qchisq(c(.9, .95, .99), df=p)
density.at.quantiles <- dchisq(x=upper.quantiles, df=p)
cut.points <- data.frame(upper.quantiles, density.at.quantiles)
dm_plot <- ggplot(data.frame(dM), aes(x=dM)) +
  geom_histogram(aes(y=after_stat(density)), bins=nclass.FD(dM),
                 fill="white", col="black") +
  geom_rug() +
  stat_function(fun=dchisq, args = list(df=p),
                col="red", linewidth=2, alpha=.7, xlim=c(0,25)) +
  geom_segment(data=cut.points,
               aes(x=upper.quantiles, xend=upper.quantiles,
                   y=rep(0,3), yend=density.at.quantiles),
               col="blue", linewidth=2) +
  xlab("Squared Mahalanobis distances and cut points") +
  ylab("Histogram and density")

Table1_elements_log$dM <- dM
Table1_elements_log$suprise <- cut(Table1_elements_log$dM,
                                   breaks = c(0, upper.quantiles, Inf),
                                   labels=c("typical", "somewhat", "Suprising", "very"))

dm_results <- data.frame(
  Sample = Table1_log[[1]],
  dM = dM,
  surprise = Table1_elements_log$suprise
)

dm_results <- dm_results %>% 
  dplyr::arrange(desc(dM)) %>% 
  head(5)
dm_plot
table(Table1_elements_log$suprise)
dm_results
```
The squared Mahalanobis distances were used to identify samples with unusual combinations of elemental concentrations. Most observations have relatively smaller Mahalanobis distance, while a smaller number have larger distance and may be considered potentially unusual. The 90%, 95%, and 99% chi-square cut-offs provide increasing levels of evidence for identifying unusual multivariate observations. These samples are retained in the analysis because an unusual observation is not necessarily an error; it may represent genuine differences in mural composition.

```{r}
#| echo: false
#| message: false
#| warning: false
#Had problems with the pairs plot from lectures mainly with the contours and scatter-points
#Two fix I built the two types of plots seperately so I can change the sizes and bins
#Then added them in seperately in the ggpairs command like done in the lectures
#The main problem was that they become squished when using the inbuilt command due to
#their being 17 variables. So doing them seperately helped fix that

surprise_points <- function(data, mapping, ...){
  ggplot(data = data, mapping = mapping) +
    geom_point(
      aes(colour = suprise),
      alpha = 0.5)}
upper_contour <- function(data, mapping, ...) {
  ggplot(data = data, mapping = mapping) +
    stat_density_2d(
      colour = "blue",
      linewidth = 0.3,
      bins = 3)}

pairs <- ggpairs(data = Table1_elements_log,
  columns = 1:17,
  lower = list(continuous = surprise_points),
  upper = list(continuous = upper_contour),
  diag = list(continuous = wrap(
      "densityDiag",
      colour = "black",
      fill = "grey"))) +
  scale_color_manual(
    values = c(
      "lightgray",
      "green",
      "blue",
      "red")) +
  theme(axis.text = element_blank(),
    axis.ticks = element_blank(),
    strip.text = element_text(size = 7))

# Legend for the pairs plot 
#Tried adding the main part to the pairs plot code itself but wouldnt show so 
#doing it seperately using cowplot. 
legend_plot <- ggplot(
  Table1_elements_log,
  aes(x = suprise, y = 1, colour = suprise)
) +
  geom_point() +
  scale_color_manual(
    values = c(
      "typical" = "lightgray",
      "somewhat" = "green",
      "Suprising" = "blue",
      "very" = "red"
    ),
    labels = c(
      "typical" = "Typical (<90%)",
      "somewhat" = "Somewhat surprising (90–95%)",
      "Suprising" = "Surprising (95–99%)",
      "very" = "Very surprising (>99%)"
    ),
    name = "Mahalanobis classification",
    drop = FALSE
  ) +
  guides(
    colour = guide_legend(
      nrow = 1,
      override.aes = list(alpha = 1, size = 3)
    )
  ) +
  theme_void() +
  theme(
    legend.position = "right",
    legend.box.margin = margin(5,10,5,10)
  )
#cowplot adds the legend that we made separately to the pairs plot
#Very annoying work but none the less like the result
legend <- cowplot::get_legend(legend_plot)
#Something that cowplot can use as ggpairs isnt a ggplot2 ggplot
pairs_grob <- GGally::ggmatrix_gtable(pairs)

final_pairs <- cowplot::plot_grid(
  pairs_grob,
  legend,
  ncol = 1,
  rel_heights = c(1, 0.14)
)
#Takes about 15 seconds to load.
final_pairs
```
The pairs plot shows the relationships among the 17 log-transformed elements. Most samples are classified as typical or somewhat surprising based on their Mahalanobis distances, while only a few are classified as more unusual. The direction and strength of the relationships vary between pairs of elements. These patterns help identify potentially unusual samples and highlight element pairs that are useful for further investigation.

## Can you assume that the data follows a multivariate normal distribution?

```{r}
#| echo: false
#We use the DM^2 values and compare them to the chi-square quantiles.
qq_elements <- Table1_elements_log[, -c(18, 19)]
n <- nrow(qq_elements)
p <- ncol(qq_elements)

chi_quantiles <- qchisq(
  ppoints(n),
  df = p
)

qq_data <- data.frame(
  Theoretical = qchisq(ppoints(n), df = p),
  Observed = sort(dM)
)

#Chi-square QQ-plot
normality_plot <- ggplot(qq_data,aes(x = Theoretical, y = Observed)) +
  geom_point(size = 2) +
  geom_abline(intercept = 0,
    slope = 1,
    linetype = "dashed",
    color = "red") +
  labs(x = "Theoretical chi-square(17) quantiles",
    y = "Ordered dm^2",
    title = "Multivariate Q-Q Plot"
  ) +
  theme_minimal()
normality_plot

#Mardia Multivariate Normal Test and individual anderson-darling tests of normality
mvn_result <- mvn(
  data = qq_elements,
  mvn_test = "mardia"
)
mvn_result$multivariate_normality
mvn_result$univariate_normality
```
The multivariate normality assumption is not fully satisfied. Mardia's skewness test is significant (p = 0.003), providing evidence against multivariate normality, although Mardia's kurtosis test is not significant (p = 0.664). The multivariate Q-Q plot is approximately linear through much of the central region but shows some departures from the reference line. Therefore, results that rely on multivariate normality, particularly the chi-square interpretation of Mahalanobis distances, should be interpreted with caution.
## Can we find groups in this data

```{r}
#| echo: false
#Correlation Heatmap: Pablo
#Fixed Issue with plot, changed to [,-1] as [,-(1:2)] includes columns 1 and 2, which removes 11-Na* from the plot
#[,-1] removes only the elements column as intended
ggcorrplot(cor(Table1_log[,-1]),
           method = "circle",
           hc.order = TRUE,
           type = "lower")
```

The correlogram shows the direction and strength of the relationships among the log-transformed elemental concentrations. Red circles represent positive correlations and blue circles represent negative correlations, while larger and darker circles indicate stronger relationships. Several element pairs show clear associations, suggesting that some elemental concentrations tend to vary together across the mural samples.

To examine these patterns more closely, two contrasting relationships from the correlogram are considered below.

### Relationship between 12-Mg\* and 14-Si\*

```{r}
#| include: false
#Look at two specific relationships too show off and talk about. 
mu <- colMeans(Table1_log[,-1])
Sigma <- var(Table1_log[,-1])
rho <- cor(Table1_log[,-1])

#Means
(mu_mg_si <- mu[c("12-Mg*", "14-Si*")])
#Covariance Matrix
(Sigma_mg_si <- Sigma[c("12-Mg*", "14-Si*"), c("12-Mg*", "14-Si*")])
#Correlation Matrix
(rho_mg_si <- rho[c("12-Mg*", "14-Si*"), c("12-Mg*", "14-Si*")])
```

```{r}
#| echo: false
#| warning: false
#| fig-width: 6
#| fig-height: 6
#| fig-align: center
#Plot of 
n <- 100
x <- seq(mu_mg_si[1]-4*sqrt(Sigma_mg_si[1,1]),
         mu_mg_si[1]+4*sqrt(Sigma_mg_si[1,1]), len=n)
y <- seq(mu_mg_si[2]-4*sqrt(Sigma_mg_si[2,2]), 
         mu_mg_si[2]+4*sqrt(Sigma_mg_si[2,2]), len=n)
xy <- expand.grid(x=x, y=y)

q.samp <- cbind(xy, prob=dmvnorm(xy, mean = mu_mg_si, sigma = Sigma_mg_si))

ggplot(q.samp, aes(x=y, y=x, z=prob)) + 
  geom_point(data=Table1_log, aes(y=`12-Mg*`, x=`14-Si*`, z=0)) +
  geom_contour(bins=5, col="red") +
  ylab("12-Mg*") + xlab("14-Si*") + ggtitle("Bivariate Scatterplot of 12-Mg* and 14-Si*", "With Two-Dimensional Density Contours") +
  coord_fixed(
    ylim = c(mu_mg_si[1]-4*sqrt(Sigma_mg_si[1,1]), mu_mg_si[1]+4*sqrt(Sigma_mg_si[1,1])),
    xlim = c(mu_mg_si[2]-4*sqrt(Sigma_mg_si[2,2]), mu_mg_si[2]+4*sqrt(Sigma_mg_si[2,2]))
    ) +
  theme_minimal()
```

The bivariate scatterplot shows a clear positive association between the log-transformed concentrations of Mg and Si. The upward-sloping density contours support this pattern: samples with higher Si concentrations generally also have higher Mg concentrations. The relatively concentrated contours suggest that this relationship is fairly strong compared with some of the other element pairs.

### Relationship between 22-Ti and 17-CI\*

```{r}
#| include: false
mu <- colMeans(Table1_log[,-1])
Sigma <- var(Table1_log[,-1])
rho <- cor(Table1_log[,-1])

#Means
(mu_mg_si2 <- mu[c("22-Ti", "17-Cl*")])
#Covariance Matrix
(Sigma_mg_si2 <- Sigma[c("22-Ti", "17-Cl*"), c("22-Ti", "17-Cl*")])
#Correlation Matrix
(rho_mg_si2 <- rho[c("22-Ti", "17-Cl*"), c("22-Ti", "17-Cl*")])
```

```{r}
#| echo: false
#| warning: false
#| fig-width: 6
#| fig-height: 6
#| fig-align: center
#Plot of 
n <- 100
x <- seq(mu_mg_si2[1]-4*sqrt(Sigma_mg_si2[1,1]),
         mu_mg_si2[1]+4*sqrt(Sigma_mg_si2[1,1]), len=n)
y <- seq(mu_mg_si2[2]-4*sqrt(Sigma_mg_si2[2,2]), 
         mu_mg_si2[2]+4*sqrt(Sigma_mg_si2[2,2]), len=n)
xy <- expand.grid(x=x, y=y)

q.samp <- cbind(xy, prob=dmvnorm(xy, mean = mu_mg_si2, sigma = Sigma_mg_si2))

ggplot(q.samp, aes(x=y, y=x, z=prob)) + 
  geom_point(data=Table1_log, aes(y=`22-Ti`, x=`17-Cl*`, z=0)) +
  geom_contour(bins=5, col="red") +
  ylab("22-Ti") + xlab("17-Cl*") + ggtitle("Bivariate Scatterplot of 22-Ti and 17-Cl*", "With Two-Dimensional Density Contours") +
  coord_fixed(
    ylim = c(mu_mg_si2[1]-4*sqrt(Sigma_mg_si2[1,1]), mu_mg_si2[1]+4*sqrt(Sigma_mg_si2[1,1])),
    xlim = c(mu_mg_si2[2]-4*sqrt(Sigma_mg_si2[2,2]), mu_mg_si2[2]+4*sqrt(Sigma_mg_si2[2,2]))
    ) +
  theme_minimal()
```

As we can see from the Bivariate Scatter plot of 22-Ti and 17-Cl, there is a negative association between the log-transformed concentrations of the two elements. the downward oreientation of the density contours suggest that samples with higher Cl concentrations generally have lower Ti concentrations. However, this relationship is relatively weaker than the relationship shown between 12-Mg and 14-Si as indicated by the dispersion of points and broader contour lines.

## PCA

We can now conduct a Principle component analysis of the data to find how many groups are in the data. The Covariance matrix was used instead of the correlation matrix to keep inline with the original studies PCA.

```{r}
#| echo: false
#PCA
PCA_log <- prcomp(Table1_log[,-1], 
                  center=TRUE, 
                  scale=FALSE)
plot(PCA_log, type="l")

PCA_variance <- data.frame(
  PC = paste0("PC", 1:length(PCA_log$sdev)),
  Eigenvalue = round(PCA_log$sdev^2, 3),
  `Proportion of Variance (%)` =
    round(100 * PCA_log$sdev^2 / sum(PCA_log$sdev^2), 2),
  `Cumulative Variance (%)` =
    round(100 * cumsum(PCA_log$sdev^2 / sum(PCA_log$sdev^2)), 2)
)

knitr::kable(
  PCA_variance,
  caption = "Variance explained by each principal component"
)
```

The first three principal components explained 77.08% of the total variance, with PC1, PC2, and PC3 accounting for 34.02%, 25.00%, and 18.05%, respectively. These results are broadly consistent with the original study, which reported 34.88%, 24.95%, and 11.32% for the first three components, and 71.15% of total variance (Sanchez del Rio et al, 2006). The first two components are particularly similar. However the third principle component in our analysis is noticeably larger then the percentage of variance in the original study.

## Can we reproduce Fig.2

```{r}
#| echo: false
#| message: false
#| warning: false
# Reproduce Figure 2: PC2 vs PC1 coloured by historical group

pca_scores <- as.data.frame(PCA_log$x[, 1:2])
pca_scores$Sample <- Table1_log[[1]]

pca_scores$Group <- case_when(
  grepl("^[Cc]ec", pca_scores$Sample)               ~ "CEC",
  grepl("^(CA|Ca|TAM|TAJ|TAC)", pca_scores$Sample)  ~ "Prehispanic",
  grepl("^(Jiu|Tot|Tez|Cua)", pca_scores$Sample)    ~ "Convents",
  TRUE                                               ~ "Other"
)

ggplot(pca_scores, aes(x = PC1, y = PC2, colour = Group, label = Sample)) +
  geom_point(size = 2.5) +
  geom_text(size = 2.2, vjust = -0.7, hjust = 0.5) +
  scale_colour_manual(values = c(
    "CEC"        = "#E41A1C",
    "Prehispanic" = "#377EB8",
    "Convents"   = "#4DAF4A"
  )) +
  labs(
    title   = "Reproduction of Figure 2",
    subtitle = "PC2 vs PC1 coloured by historical group",
    x = "PC 1",
    y = "PC 2",
    colour = "Group"
  ) +
  theme_bw()
```

The PCA plot broadly reproduces Figure 2 from Sánchez del Río et al. (2006). The CEC samples (Santa Cecilia Acatitlán) form a tight, well-separated cluster, consistent with their distinct non-fresco preparation technique. The Prehispanic and Convent groups occupy partially overlapping regions of the PC1–PC2 plane, suggesting that, despite their temporal separation, both groups employed similar mural painting techniques. It is also worth noting that we can observe that despite being listed in the Mexican Franciscan convents group, both Cuauhtinchan samples cluster around the Prehispanic murals, suggesting its work is more closely related to this type of mural. This was also seen in the original papers figure, however there is more noticeable overlap with this figure compared to the original. 


# Discussion

Answering the questions from the document and any other conclusions we have from the analysis

1.  Can you find groups in these data?
2.  Can you reproduce Fig. 2? Do you agree with the conclusions the authors draw about the “zones”?

# References {style="text-align: left;"}

> Andonie, R., and I. Dzitac. 2010. “How to Write a Good Paper in Computer Science and       How Will It Be Measured by ISI Web of Knowledge.” International Journal of Computers,     Communications & Control V (4): 432–44. https://doi.org/10.15837/ijccc.2010.4.2493.

> Frery, A. C., L. Gomez, and A. C. Medeiros. 2020. “A Badging System for Reproducibility     and Replicability in Remote Sensing Research.” IEEE Journal of Selected Topics on        Applied Earth Observations and Remote Sensing 13: 4988–95.                               https://doi.org/10.1109/JSTARS.2020.3019 418.

> Sánchez del Río, M., Martinetto, P., Solís, C. & Reyes-Valerio, C. (2006), ‘PIXE           analysis on Maya blue in Prehispanic and colonial mural paintings’, Nuclear              Instruments and Methods in Physics Research Section B: Beam Interactions with            Materials and Atoms 249(1–2), 628–632.
