---
title: "Group 8 Project"
subtitle: "PIXE analysis on Maya blue in Prehispanic and colonial mural paintings"
author: "Awet, Connor, Lachlan, Pablo,"
format: 
  html:
    toc: true
    toc-depth: 3
---

```{r}
#| include: false
library("readxl")
library(tidyverse)
library(gt)
library(GGally)
library(cowplot)
library(MVN)
library(ggcorrplot)
library(mvtnorm)
Table1_raw <- read_excel("../Data/Table1_raw.xlsx")

Table1 <- read_excel("../Data/Table1_clean.xlsx")
Table1$`17-Cl*` <- as.numeric(Table1$`17-Cl*`)
Table1$`15-P` <- as.numeric(Table1$`15-P`)
Table1$`29-Cu` <- as.numeric(Table1$`29-Cu`)
```

# Introduction

The original article, PIXE analysis on Maya blue in Prehispanic and colonial mural paintings, by Sanchez del Rio et al (2006), sampled, analysed and grouped using PCA on murals containing Maya Blue across different archaeological sites. Across the sites they collected 35 samples from murals and either the elements concentration percent or the ratio of micro gram to gram ratio of that element in the mural. From these murals they sampled 17 different chemical elements from sodium to lead. They found with their PCA that 70% of the total variance in the sample population can be described using only the first three principle components and was similar to what was observed in other art forms. The three zones found were CEC, Prehispanic, and convents. with the first clustering in a small group separate from the rest of the samples.

The aim of this analysis is too find groups within the data and reproduce what was found in the original paper. The reason for this is to support the conclusions made by Sanchez del rio et al (2006) about the 3 zones found in the data during their analysis. By doing this we can reassert the claims made about the differences between the different types of murals and their distinct zones of style. This report will contain both a quantitative and qualitative analysis of the data, as well as PCA and cluster analysis to explore the possible groups in the data.

# Methodology

## Data Transformation

The data was transformed in line with the original study to be able to address different issues with comparison and estimating of element concentrations. Different concentration ranges: The original data contain elemental concentrations reported either as percentages or as µg/g (micrograms per gram). Because these measurements have substantially different concentration ranges, a (\log\_{10}) transformation is applied to the concentrations. This follows the procedure used in the original paper, where the authors state that the logarithm of the concentrations was used to avoid large discrepancies in concentration range. Values below the detection limit: Some concentrations in Table 1 are reported in parentheses ( ). According to the original paper, these values are below the limit of detection (LOD). Since numerical estimates are still provided, these estimated concentrations are retained for the analysis and the parentheses are removed. Lastly we conducted a transformation of both NA and 0 values found in the data. Sanchez del Rio et al (2006) described that values that were given as 0's arent treated as 0's, and were instead concentrations that were below the detection limit, and as such were instead very small concentrations of elements. Moreover we cannot do a (\log\_{10}) transformation of the data if observations contain 0's. This leaves us to therefore transform these 0's into estimates to both do this transformation, and not remove concentrations that are there but very small. Therefore we run a function that runs through each column that is numeric, and then filters out any values that are less than or equal to zero, ignores NAs (na.rm = TRUE), finds the smallest remaining positive number, and divides it by two. Finally it checks every cell in that column. If a cell is either NA or exactly 0, it substitutes it with the replacement value.

# Results

<!-- -->

## Quantitative Descriptive Analysis

```{r}
#| include: false
### Kept this transformation in the method section.

#Non transformed table - dont use. 
# Make estimated concentrations positive
Table1 <- Table1 %>% 
  mutate(across(where(is.numeric), abs))

# Replace values recorded as zero with half of the
# smallest positive value for that element
Table1 <- Table1 %>%
  mutate(across(
    where(is.numeric),
    ~ replace(
        .x,
        is.na(.x) | .x == 0,
        min(.x[.x > 0], na.rm = TRUE) / 2
      )
  ))

#Log Table - Use as variables become comparable
Table1_log <- Table1 %>% 
  mutate(across(where(is.numeric), log10))

#Change table around to long ways for boxplots in future as current table arrangement didnt work.

#This one incase we need it for the non-transformed. 
Table1_long <- Table1 %>%
  pivot_longer(
    cols = -Elements,
    names_to = "Element_Group",     
    values_to = "Concentrations"   
  )

#Main Table we use for boxplot
Table1_long_log <- Table1_log %>%
  pivot_longer(
    cols = -Elements,
    names_to = "Element_Group",     
    values_to = "Concentrations"
  )
```

```{r}
#| echo: false
# Descriptive table of Raw data - Show why we need to transform
desc_pixe <- Table1 %>% 
  summarise(
    across(
      where(is.numeric),
      list(
        M = ~mean(.x, na.rm = TRUE),
        SD = ~sd(.x, na.rm = TRUE),
        Median = ~median(.x, na.rm = TRUE),
        Min = ~min(.x, na.rm =TRUE),
        Q1 = ~round(quantile(.x, probs = 0.25, na.rm = TRUE),2),
        IQR = ~round((quantile(.x, probs = 0.75, na.rm =TRUE) - quantile(.x, probs = 0.25, na.rm = TRUE)), 2),
        Q3 = ~round(quantile(.x, probs = 0.75, na.rm = TRUE),2),
        Max = ~max(.x, na.rm =TRUE)
      )
    )
  ) %>% 
  pivot_longer(
    everything(),
    names_to = c("Elements", "Statistics"),
    names_sep = "_",
    values_to = "Value"
  ) %>% 
  pivot_wider(
    names_from = Statistics,
    values_from = Value
  )

output_desc_pix <- desc_pixe %>% 
  gt() %>% 
  cols_label(
    Elements = "Element",
    M = md("Mean"),
    SD = md("Standard D."),
    Median = "Median",
    Min = "Minimum",
    Q1 = "1st Quartile",
    IQR = "Inter-Q Range",
    Q3 = "3rd Quartile",
    Max = "Maximum"
  ) %>% 
  fmt_number(
    columns = c(M, SD, Median, Min, Max),
    decimals = 2
  ) %>% 
  tab_header(
    title = "Descriptive Statistics of Elemental Concentrations",
    subtitle = "Non-Transformed"
  ) %>% 
  tab_options(
    table.font.size = 12
  )
output_desc_pix
gtsave(output_desc_pix, "output_desc_pix.png")
```
The dataset is highly heterogeneous. Calcium is a dominant major element in most samples, while several trace elements span orders of magnitude. This is exactly why the authors analysed logarithms of concentration. Two especially large phosphorus values occur for Totimehuacan: 99,600 $\mu$g/g in `Tot1_1` and 51,800 $\mu$g/g in `Tot1_3`. These values are retained because they are printed in Table 1; the source does not provide enough evidence to assign a specific contamination or measurement-error cause to them.
```{r}
#| echo: false
# Log Descriptives Table
desc_pixe_log <- Table1_log %>% 
  summarise(
    across(
      where(is.numeric),
      list(
        M = ~mean(.x, na.rm = TRUE),
        SD = ~sd(.x, na.rm = TRUE),
        Median = ~median(.x, na.rm = TRUE),
        Min = ~min(.x, na.rm =TRUE),
        Q1 = ~round(quantile(.x, probs = 0.25, na.rm = TRUE),2),
        IQR = ~round((quantile(.x, probs = 0.75, na.rm =TRUE) - quantile(.x, probs = 0.25, na.rm = TRUE)), 2),
        Q3 = ~round(quantile(.x, probs = 0.75, na.rm = TRUE),2),
        Max = ~max(.x, na.rm =TRUE)
      )
    )
  ) %>% 
  pivot_longer(
    everything(),
    names_to = c("Elements", "Statistics"),
    names_sep = "_",
    values_to = "Value"
  ) %>% 
  pivot_wider(
    names_from = Statistics,
    values_from = Value
  )
output_desc_pix_log <- desc_pixe_log %>% 
  gt() %>% 
  cols_label(
    Elements = "Element",
    M = md("Mean"),
    SD = md("Standard D."),
    Median = "Median",
    Min = "Minimum",
    Q1 = "1st Quartile",
    IQR = "Inter-Q Range",
    Q3 = "3rd Quartile",
    Max = "Maximum"
  ) %>% 
  fmt_number(
    columns = c(M, SD, Median, Min, Max),
    decimals = 2
  ) %>% 
  tab_header(
    title = "Descriptive Statistics of Elemental Concentrations",
    subtitle = "Log10 Transformation"
  ) %>% 
  tab_options(
    table.font.size = 12
  )
output_desc_pix_log
```

## Graphical and Qualitative Analysis

```{r}
#| echo: false
#| message: false
#| warning: false
#Box Plot of transformed data
decp_boxplot <- ggplot(data = Table1_long_log, aes(x = Element_Group, y = Concentrations)) + 
  geom_boxplot(aes(fill = Element_Group), notch = TRUE, outlier.shape = 1.5) + 
  theme_bw() +
  theme(axis.text.x = element_text(size = 7, angle = 90, hjust = 1, vjust = 0.5), 
        legend.position = "none")
decp_boxplot
```
The log-scale boxplots show that transformation reduces, but does not remove, the pronounced differences in spread among elements. The trace variables remain particularly heterogeneous, while several major matrix components are more concentrated.

The calcium plot provides a stronger archaeological interpretation than a purely generic distribution plot. @sanchez2006pixe report that most mural samples have Ca concentrations around 30--50%, close to the 40% Ca content of calcite, supporting use of the fresco technique. The CEC samples are the conspicuous exception, with substantially lower Ca, consistent with the authors' conclusion that these murals were not prepared by the same fresco process.
## Are There Surprising Observations?

```{r}
#| echo: false
# Mahalanobis Distances 
Table1_elements_log <- Table1_log[, -1]
mu_hat <- colMeans(Table1_elements_log)
Sigma_hat <- cov(Table1_elements_log)
p <- ncol(Table1_elements_log)

dM <- mahalanobis(Table1_elements_log, center=mu_hat, cov=Sigma_hat)
upper.quantiles <- qchisq(c(.9, .95, .99), df=p)
density.at.quantiles <- dchisq(x=upper.quantiles, df=p)
cut.points <- data.frame(upper.quantiles, density.at.quantiles)
dm_plot <- ggplot(data.frame(dM), aes(x=dM)) +
  geom_histogram(aes(y=after_stat(density)), bins=nclass.FD(dM),
                 fill="white", col="black") +
  geom_rug() +
  stat_function(fun=dchisq, args = list(df=p),
                col="red", linewidth=2, alpha=.7, xlim=c(0,25)) +
  geom_segment(data=cut.points,
               aes(x=upper.quantiles, xend=upper.quantiles,
                   y=rep(0,3), yend=density.at.quantiles),
               col="blue", linewidth=2) +
  xlab("Squared Mahalanobis distances and cut points") +
  ylab("Histogram and density")

Table1_elements_log$dM <- dM
Table1_elements_log$suprise <- cut(Table1_elements_log$dM,
                                   breaks = c(0, upper.quantiles, Inf),
                                   labels=c("typical", "somewhat", "Suprising", "very"))

dm_results <- data.frame(
  Sample = Table1_log[[1]],
  dM = dM,
  surprise = Table1_elements_log$suprise
)

dm_results <- dm_results %>% 
  dplyr::arrange(desc(dM)) %>% 
  head(5)
dm_plot
table(Table1_elements_log$suprise)
dm_results
```

```{r}
#| echo: false
#| message: false
#| warning: false
#Had problems with the pairs plot from lectures mainly with the contours and scatter-points
#Two fix I built the two types of plots seperately so I can change the sizes and bins
#Then added them in seperately in the ggpairs command like done in the lectures
#The main problem was that they become squished when using the inbuilt command due to
#their being 17 variables. So doing them seperately helped fix that

surprise_points <- function(data, mapping, ...){
  ggplot(data = data, mapping = mapping) +
    geom_point(
      aes(colour = suprise),
      alpha = 0.5)}
upper_contour <- function(data, mapping, ...) {
  ggplot(data = data, mapping = mapping) +
    stat_density_2d(
      colour = "blue",
      linewidth = 0.3,
      bins = 3)}

pairs <- ggpairs(data = Table1_elements_log,
  columns = 1:17,
  lower = list(continuous = surprise_points),
  upper = list(continuous = upper_contour),
  diag = list(continuous = wrap(
      "densityDiag",
      colour = "black",
      fill = "grey"))) +
  scale_color_manual(
    values = c(
      "lightgray",
      "green",
      "blue",
      "red")) +
  theme(axis.text = element_blank(),
    axis.ticks = element_blank(),
    strip.text = element_text(size = 7))

# Legend for the pairs plot 
#Tried adding the main part to the pairs plot code itself but wouldnt show so 
#doing it seperately using cowplot. 
legend_plot <- ggplot(
  Table1_elements_log,
  aes(x = suprise, y = 1, colour = suprise)
) +
  geom_point() +
  scale_color_manual(
    values = c(
      "typical" = "lightgray",
      "somewhat" = "green",
      "Suprising" = "blue",
      "very" = "red"
    ),
    labels = c(
      "typical" = "Typical (<90%)",
      "somewhat" = "Somewhat surprising (90–95%)",
      "Suprising" = "Surprising (95–99%)",
      "very" = "Very surprising (>99%)"
    ),
    name = "Mahalanobis classification",
    drop = FALSE
  ) +
  guides(
    colour = guide_legend(
      nrow = 1,
      override.aes = list(alpha = 1, size = 3)
    )
  ) +
  theme_void() +
  theme(
    legend.position = "right",
    legend.box.margin = margin(5,10,5,10)
  )
#cowplot adds the legend that we made separately to the pairs plot
#Very annoying work but none the less like the result
legend <- cowplot::get_legend(legend_plot)
#Something that cowplot can use as ggpairs isnt a ggplot2 ggplot
pairs_grob <- GGally::ggmatrix_gtable(pairs)

final_pairs <- cowplot::plot_grid(
  pairs_grob,
  legend,
  ncol = 1,
  rel_heights = c(1, 0.14)
)
#Takes about 15 seconds to load.
final_pairs
```

## Can you assume that the data follows a multivariate normal distribution?

```{r}
#| echo: false
#We use the DM^2 values and compare them to the chi-square quantiles.
qq_elements <- Table1_elements_log[, -c(18, 19)]
n <- nrow(qq_elements)
p <- ncol(qq_elements)

chi_quantiles <- qchisq(
  ppoints(n),
  df = p
)

qq_data <- data.frame(
  Theoretical = qchisq(ppoints(n), df = p),
  Observed = sort(dM)
)

#Chi-square QQ-plot
normality_plot <- ggplot(qq_data,aes(x = Theoretical, y = Observed)) +
  geom_point(size = 2) +
  geom_abline(intercept = 0,
    slope = 1,
    linetype = "dashed",
    color = "red") +
  labs(x = "Theoretical chi-square(17) quantiles",
    y = "Ordered dm^2",
    title = "Multivariate Q-Q Plot"
  ) +
  theme_minimal()
normality_plot

#Mardia Multivariate Normal Test and individual anderson-darling tests of normality
mvn_result <- mvn(
  data = qq_elements,
  mvn_test = "mardia"
)
mvn_result$multivariate_normality
mvn_result$univariate_normality
```

## Can we find groups in this data

```{r}
#| echo: false
#Correlation Heatmap: Pablo
#Fixed Issue with plot, changed to [,-1] as [,-(1:2)] includes columns 1 and 2, which removes 11-Na* from the plot
#[,-1] removes only the elements column as intended
ggcorrplot(cor(Table1_log[,-1]),
           method = "circle",
           hc.order = TRUE,
           type = "lower")
```

\[need to explain correlogram here\]

From the correlogram we can pick two examples of the relationships between elements existing in the data

### Relationship between 12-Mg\* and 14-Si\*

```{r}
#| include: false
#Look at two specific relationships too show off and talk about. 
mu <- colMeans(Table1_log[,-1])
Sigma <- var(Table1_log[,-1])
rho <- cor(Table1_log[,-1])

#Means
(mu_mg_si <- mu[c("12-Mg*", "14-Si*")])
#Covariance Matrix
(Sigma_mg_si <- Sigma[c("12-Mg*", "14-Si*"), c("12-Mg*", "14-Si*")])
#Correlation Matrix
(rho_mg_si <- rho[c("12-Mg*", "14-Si*"), c("12-Mg*", "14-Si*")])
```

```{r}
#| echo: false
#| warning: false
#| fig-width: 6
#| fig-height: 6
#| fig-align: center
#Plot of 
n <- 100
x <- seq(mu_mg_si[1]-4*sqrt(Sigma_mg_si[1,1]),
         mu_mg_si[1]+4*sqrt(Sigma_mg_si[1,1]), len=n)
y <- seq(mu_mg_si[2]-4*sqrt(Sigma_mg_si[2,2]), 
         mu_mg_si[2]+4*sqrt(Sigma_mg_si[2,2]), len=n)
xy <- expand.grid(x=x, y=y)

q.samp <- cbind(xy, prob=dmvnorm(xy, mean = mu_mg_si, sigma = Sigma_mg_si))

ggplot(q.samp, aes(x=y, y=x, z=prob)) + 
  geom_point(data=Table1_log, aes(y=`12-Mg*`, x=`14-Si*`, z=0)) +
  geom_contour(bins=5, col="red") +
  ylab("12-Mg*") + xlab("14-Si*") + ggtitle("Bivariate Scatterplot of 12-Mg* and 14-Si*", "With Two-Dimensional Density Contours") +
  coord_fixed(
    ylim = c(mu_mg_si[1]-4*sqrt(Sigma_mg_si[1,1]), mu_mg_si[1]+4*sqrt(Sigma_mg_si[1,1])),
    xlim = c(mu_mg_si[2]-4*sqrt(Sigma_mg_si[2,2]), mu_mg_si[2]+4*sqrt(Sigma_mg_si[2,2]))
    ) +
  theme(axis.text.x = element_text(angle = 45))
```

We can see from the bivariate scatter plot that their is a clear positive relationship between 12-Mg and 14-Si. As we can see with the upward-sloping denisty contour lines, the when the log-transformed concentration of Si increases, the log-transformed concentration of Mg also tends to increase.

### Relationship between 22-Ti and 17-CI\*

```{r}
#| include: false
mu <- colMeans(Table1_log[,-1])
Sigma <- var(Table1_log[,-1])
rho <- cor(Table1_log[,-1])

#Means
(mu_mg_si <- mu[c("22-Ti", "17-Cl*")])
#Covariance Matrix
(Sigma_mg_si <- Sigma[c("22-Ti", "17-Cl*"), c("22-Ti", "17-Cl*")])
#Correlation Matrix
(rho_mg_si <- rho[c("22-Ti", "17-Cl*"), c("22-Ti", "17-Cl*")])
```

```{r}
#| echo: false
#| warning: false
#| fig-width: 6
#| fig-height: 6
#| fig-align: center
#Plot of 
n <- 100
x <- seq(mu_mg_si[1]-4*sqrt(Sigma_mg_si[1,1]),
         mu_mg_si[1]+4*sqrt(Sigma_mg_si[1,1]), len=n)
y <- seq(mu_mg_si[2]-4*sqrt(Sigma_mg_si[2,2]), 
         mu_mg_si[2]+4*sqrt(Sigma_mg_si[2,2]), len=n)
xy <- expand.grid(x=x, y=y)

q.samp <- cbind(xy, prob=dmvnorm(xy, mean = mu_mg_si, sigma = Sigma_mg_si))

ggplot(q.samp, aes(x=y, y=x, z=prob)) + 
  geom_point(data=Table1_log, aes(y=`22-Ti`, x=`17-Cl*`, z=0)) +
  geom_contour(bins=5, col="red") +
  ylab("22-Ti") + xlab("17-Cl*") + ggtitle("Bivariate Scatterplot of 22-Ti and 17-Cl*", "With Two-Dimensional Density Contours") +
  coord_fixed(
    ylim = c(mu_mg_si[1]-4*sqrt(Sigma_mg_si[1,1]), mu_mg_si[1]+4*sqrt(Sigma_mg_si[1,1])),
    xlim = c(mu_mg_si[2]-4*sqrt(Sigma_mg_si[2,2]), mu_mg_si[2]+4*sqrt(Sigma_mg_si[2,2]))
    ) +
  theme(axis.text.x = element_text(angle = 45))
```

As we can see from the Bivariate Scatter plot of 22-Ti and 17-Cl, there is a negative association between the log-transformed concentrations of the two elements. the downward oreientation of the density contours suggest that samples with higher Cl concentrations generally have lower Ti concentrations. However, this relationship is relatively weaker then the relationship shown between 12-Mg and 14-Si as indicated by the dispersion of points and broader contour lines.

##PCA

We can now conduct a Principle component analysis of the data to find how many groups are in the data. The Covariance matrix was used instead of the correlation matrix to keep inline with the original studies PCA.

```{r}
#| echo: false
#PCA
PCA_log <- prcomp(Table1_log[,-1], 
                  center=TRUE, 
                  scale=FALSE)
plot(PCA_log, type="l")

PCA_variance <- data.frame(
  PC = paste0("PC", 1:length(PCA_log$sdev)),
  Eigenvalue = round(PCA_log$sdev^2, 3),
  `Proportion of Variance (%)` =
    round(100 * PCA_log$sdev^2 / sum(PCA_log$sdev^2), 2),
  `Cumulative Variance (%)` =
    round(100 * cumsum(PCA_log$sdev^2 / sum(PCA_log$sdev^2)), 2)
)

knitr::kable(
  PCA_variance,
  caption = "Variance explained by each principal component"
)
```

The first three principal components explained 77.08% of the total variance, with PC1, PC2, and PC3 accounting for 34.02%, 25.00%, and 18.05%, respectively. These results are broadly consistent with the original study, which reported 34.88%, 24.95%, and 11.32% for the first three components (Sanchez del Rio et al, 2006). The first two components are particularly similar. However the third principle component in our analysis is noticeably larger then the percentage of variance in the original study.

## Can we reproduce Fig.2

```{r}
#| echo: false
#Cluster Analysis
```

# Discussion

Answering the questions from the document and any other conclusions we have from the analysis

1.  Can you find groups in these data?
2.  Can you reproduce Fig. 2? Do you agree with the conclusions the authors draw about the “zones”?

# References {style="text-align: center;"}

> Andonie, R., and I. Dzitac. 2010. “How to Write a Good Paper in Computer Science and How Will It Be Measured by ISI Web of Knowledge.” International Journal of Computers, Communications & Control V (4): 432–44. https://doi.org/10.15837/ijccc.2010.4.2493.
>
> Frery, A. C., L. Gomez, and A. C. Medeiros. 2020. “A Badging System for Reproducibility and Replicability in Remote Sensing Research.” IEEE Journal of Selected Topics on Applied Earth Observations and Remote Sensing 13: 4988–95. https://doi.org/10.1109/JSTARS.2020.3019 418.
>
> Sánchez del Río, M., Martinetto, P., Solís, C. & Reyes-Valerio, C. (2006), ‘PIXE analysis on Maya blue in Prehispanic and colonial mural paintings’, Nuclear Instruments and Methods in Physics Research Section B: Beam Interactions with Materials and Atoms 249(1–2), 628–632.
